### 2017-01-22 text/Chapter06.tex
Edits pages 125-187

We mirror the development for flat coeffect calculus and develop a small calculus with a type system
that captures per-variable contextual properties. We outline its categorical semantics and use it as
a basis for a translation that turns well-typed programs in context-aware languages into well-typed
programs in a simple target functional language. We prove syntactic safety for a sample target
language, showing that ``well-typed context-aware programs do not get stuck''.

@@

\item We give a \emph{translational semantics} (Section~\ref{sec:struct-transl}) that translates
  programs from the structural coeffect calculus into a simple functional language with
  uninterpreted comonadically-inspired primitives. We give concrete operational semantics for the
  target language for one of our sample languages and show that well-typed programs, produced by
  translation from the coeffect calculus do not get stuck.
\end{itemize}

@@

In the previous chapter, we discussed the correspondence between coeffects and effects (and
between comonads and monads). As noted in Section~\ref{sec:applications-strucutre-vec}, the
$\lambda$-calculus is asymmetric in that an expression has multiple inputs (variables in the
context), but just a single result (the resulting value). So, while there is only one notion of
monadic effect system, there are two separate notions of coeffect system - one that keeps coeffect
annotations per-environment and one that keeps coeffect annotations per-variable.

@@

In the structural coeffect calculus, functions are annotated with a primitive (scalar) coeffect
annotations. A vector of variables forming the free-variable context is annotated with a vector of
coeffect annotations. These annotations differ for various coeffect calculi; their properties are
captured by the definition of \emph{structural coeffect scalar} below. The scalar annotations can
be \eg~integers (how many past values we need) or values of a two-point lattice specifying whether
a variable is live or not. The expressions and types of the structural coeffect calculus are defined
as follows:
%
\begin{equation*}
\begin{array}{rcl}
e &::=& x \sep n \sep \lambda x:\tau.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
\tau &::=& \ident{num} \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
\end{array}
\end{equation*}
%
The expressions and types of structural coeffect calculus are similar to those of the flat coeffect
calculus with two differences. First, we omit the \kvd{let} construct in the core language. In
structural coeffects, let binding can be defined as a derived rule using abstraction and application
(Section~\ref{sec:structural-exts-let}). Second, the coeffect annotations $\cclrd{r}, \cclrd{s}, \cclrd{t}$
on function type now range over values of a \emph{structural coeffect scalar}.

@@

The \emph{structural coeffect scalar} structure is similar to that of \emph{flat coeffect algebra}
with the exception that it drops the $\czip$ operation. It only provides a monoid $(\C, \cseq, \cunit)$
modelling sequential composition of computations and a monoid $(\C, \cpar, \czero)$ representing
pointwise composition, as well as the $\cleq$ relation.

\begin{definition}
\label{def:structural-scalar}
A \emph{\cclrd{structural coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set
$\C$ together with elements $\cunit, \czero \in \C$, binary operations $\cseq, \cpar$ such that
$(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and a binary relation $\cleq$ such
that $(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
%
\begin{equation*}
\begin{array}{ccr}
r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
\textnormal{(monoid)}
\\
r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
\czero \;\cpar\; r = r = r \;\cpar\; \czero &
\textnormal{(monoid)}
\\
\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
t\; \cleq\; t &
\textnormal{(pre-order)}
\\
\end{array}
\end{equation*}
%
In addition, the following distributivity axioms hold:
\begin{align*}
\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
\end{align*}
\end{definition}

\noindent
The structural coeffect scalar structure resembles flat coeffect algebra, but it differs in two important ways:
%
\begin{itemize}
\item The $\cpar$ operation of structural coeffect scalar is not required to be idempotent. In structural
  systems, we can track individual variable accesses and not requiring idempotence allows interesting
  systems such as that for bounded reuse (Section~\ref{sec:struct-semantics-examples}).

@@

\noindent
In the structural coeffect calculus, the scalar coeffect structure is supplemented by a vector
structure. The vector structure is used to manipulate vectors of coeffect scalars attached to a
variable context. The required structure is captured by the following definition.

\begin{definition}
A \emph{\aclrd{structural coeffect algebra}} is formed by a structural coeffect scalar
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ equipped with the following additional structures:

\begin{itemize}
\vspace{-0.3em}
\item Coeffect vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$, ranging over
  structural coeffect scalars indexed by vector lengths $\sclrd{m}, \sclrd{n} \in \sclrd{\mathbb{N}}$.

@@

In the structural system, this is made explicit by adding \emph{structural rules}.
While substructural type systems usually remove some of the rules, we keep all three and
use them to track how variables are used. This is done by manipulating the coeffect annotations
in parallel with manipulating the variable contexts. As in substructural type systems, (\emph{app})
checks the types of sub-expressions in disjoint parts of the free variable contexts and
(\emph{var}) requires the context to contain exactly one variable. The typing rule for let binding in
structural coeffects is a derived rule and we discuss it later in Section~\ref{sec:structural-exts-let}.

@@

\paragraph{Variable contexts.}
In Chapter~\ref{ch:flat}, the free variable context $\Gamma$ was treated as a set. In the type
system for the structural coeffect calculus, the variable context is treated as a vector, with
an additional condition that a cannot appear multiple times. We also write $\xlen{-}$ for the
length of the vector:
%
\begin{equation*}
\begin{array}{l}
 \Gamma = \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle \qquad \textnormal{such that}~~\forall i, j ~.~ i \neq j \implies x_i \neq x_j\\
 \xlen{\langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle} = n
\end{array}
\end{equation*}
%
We use the usual notation $x_1\!:\!\tau_1, \ldots, x_1\!:\!\tau_1 \vdash e : \tau$ for
typing judgements, but the free variable context should be understood as a vector.
The notation $\Gamma_1, \Gamma_2$ is used for concatenation of vectors of variables. That is,
given a context $\Gamma_1 = \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle $ and a context
$\Gamma_2 = \langle x_{n+1}\!:\!\tau_{n+1}, \ldots, x_m\!:\!\tau_m \rangle $ then
$\Gamma_1, \Gamma_2 = \langle x_1\!:\!\tau_1, \ldots, x_m\!:\!\tau_m \rangle$.

In the typing rules, free variable contexts are annotated with vectors of structural coeffect scalars,
written as $\coctx{x_1\!:\tau_1,\ldtos,x_n\!:\!\tau_n}{\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}}$.
Meta-variables ranging over coeffect vectors are written as $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$
(using bold face and colour to distinguish them from scalar meta-variables) and the length of a
coeffect vector is written as $\alen{\aclrd{\textbf{r}}}$.

@@

\paragraph{Structural rules.}
These are shown in Figure~\ref{fig:struct-types} (b). The three structural rules are not
syntax-directed and allow different transformation of the free variable context. They correspond
to the transformations known as \emph{weakening}, \emph{exchange} and \emph{contraction}
from substructural systems.

@@

This suggests that $\czip$ and $\cpar$ serve two roles in flat coeffects. First, they are used
as over-approximations and under-approximations of $\atimes$. This is demonstrated by the
(\emph{approximation}) requirement introduced in Section~\ref{sec:flat-syntax-cbv}, which requires that
$\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$. Semantically, flat abstraction
combines two values representing the available context, potentially discarding parts of it
(under-approximation), while flat application splits the available context (a single value),
potentially duplicating parts of it (over-approxi\-mation)\footnote{Because of this duality, earlier
version of coeffect systems \cite{coeffects-icalp13} used $\wedge$ and $\vee$.}.

@@

\subsection{Examples of structural coeffects}

The structural coeffect calculus above can be instantiated to obtain the 3 structural coeffect calculi
presented in Section~\ref{sec:applications-structural}. Two of them -- structural dataflow and
structural liveness provide a more precise tracking of properties that can be tracked using
flat systems. Formally, any flat coeffect algebra can be turned into a structural coeffect scalar
(by dropping the $\czip$ operator). This is useful for liveness and dataflow, but it does not yield
a practically useful system for the flat algebra coeffect for implicit parameters.

On the other hand, some of the structural systems do not have a flat equivalent, typically
because there is no appropriate $\czip$ operator that could be added to form the flat coeffect
scalar. This is the case, for example, for the system tracking bounded variable use
(Example~\ref{ex:structural-ex-bll}).

\begin{example}[Structural liveness]
The structural coeffect scalar for liveness is formed by
$(\mathcal{L}, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$, where $\mathcal{L}=\{ \ident{L}, \ident{D} \}$ is
the same two-point lattice as in the flat version, that is $\ident{D} \sqsubseteq \ident{L}$
with a join $\sqcup$ and a meet $\sqcap$.
\end{example}

\begin{example}[Structural dataflow]
In dataflow, context is annotated with natural numbers and the structural coeffect scalar is formed
by $(\mathbb{N}, +, \mathit{max}, 0, 0, \leq)$.
\end{example}

\noindent
These two examples have both flat and structural versions. For them, obtaining the structural
coeffect algebra is easy. As shown by the examples above, we simply omit the $\czip$ operation. The
laws required by a structural coeffect algebra are the same as those required by the flat version
and so the above definitions are both valid. Similar construction can be used for the
\emph{optimized dataflow} example from Section~\ref{sec:flat-calculus-examples}.

It is important to note that this gives us a systems with \emph{different} properties.
Information is now tracked per-variable rather than for entire contexts. For dataflow,
we also need to adapt the typing rule for the \kvd{prev} construct. Here, we write $\aclrd{+}$
for a pointwise extension of the $+$ operator, such that
$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} \aclrd{+} k = \alift{\cclrd{r_1} + k, \ldots, \cclrd{r_n} + k}$.
%
\begin{equation*}
\tyrule{prev}
  { \coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau }
  { \coctx{\Gamma}{\aclrd{\textbf{r}} \,\aclrd{+}\, 1} \vdash \kvd{prev}~e : \tau }
\end{equation*}
%
The rule appears similar to the flat one, but there is an important difference. Because of the
structural nature of the type system, it only increments the required number of values for
variables that are actually used in the expression $e$, whereas in the flat coeffect system the rule
incremented the annotation for the whole context. Thanks to the structural nature of the system,
annotations of variables that do not appear in the expression $e$ can be left unchanged.

Before looking at properties of structural coeffect systems,
we consider a system for tracking bounded variable use, which is an example of structural system
that does not have a flat counterpart.

@@

However, the type system for structural coeffect calculus in Figure~\ref{fig:struct-types}
introduces another kind of ambiguity due to the fact that non-syntax-directed structural rules
can be applied repeatedly and in arbitrary order. As with the semantics for flat coeffect calculus
in Chapter~\ref{ch:semantics}, we define the semantics of the structural coeffect calculus relative to
a \emph{typing derivation} and so the meaning of a program depends on the typing derivation chosen.
In this section, we specify how to choose the desired \emph{unique} typing derivation, following
the example of flat coeffect calculi as discussed in Section~\ref{sec:flat-unique}.

@@

As with $\ident{findvar}_{x,\tau}$, the \ident{mergevars} function is fully deterministic.
It returns context consisting of three parts. Parts $\Gamma_1$
and $\Gamma_2$ represent variables that appear only in the first or the second context; part
$\Gamma$ contains common variables. The coeffect annotations corresponding to $\Gamma_1$ are
the original annotations from $\aclrd{\textbf{r}}$; the coeffects corresponding to $\Gamma_2$
are composed with the coeffect of the function value $\cclrd{t} \,\aseq\, \aclrd{\textbf{s}'}$
as in the original (\emph{app}) rule. Finally, for shared variables, the coeffect is
obtained by point-wise composition (as in contraction) of the coeffect for the two contexts
$\cclrd{r_i} \;\cpar\; (\cclrd{t} \,\cseq\, \cclrd{s_j})$. The first coeffect corresponds to the
context demands in the sub-expression $e_1$ and the second coeffect corresponds to the
function argument $e_2$ sequentially composed with the coeffect $\cclrd{t}$ of the function
(as in ordinary application rule).

@@

When discussing the flat coeffect calculus in Section~\ref{sec:flat-calculus-algebra}, we noted
that the $\cleq$ operation for \emph{flat coeffect algebra} can be defined in terms of $\cpar$
as follows:
%
\begin{equation*}
\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}
\end{equation*}
%
This is not the case for the \emph{structural coeffect scalar} structure. For example, in the
calculus for tracking bounded reuse, the $\cpar$ operator is defined as $+$ (on integers) and
$\cleq$ is just $\leq$ and so the above equivalence does not hold. For this reason, we included
$\cleq$ as an explicit part of both of the structures.

@@

The properties of the structural coeffect algebra, together with two additional weak conditions,
guarantee that certain equational properties on terms hold in all instances of the structural
coeffect calculus that we consider in this thesis. In this section, we look at these common
properties. In Section~\ref{sec:struct-syntax-intro}, we first briefly compare equational theory
for flat coeffects (Section~\ref{sec:flat-syntax}) and structural coeffects (Section~\ref{sec:struct-syntactic-thms}).

@@

The following example shows a situation similar to the previous one. Here, we use
structural dataflow calculus (writing $\kvd{prev}~e$ to obtain previous value of the
expression $e$) and we substitute $w+z$ for $y$:
%
\begin{equation*}
\begin{array}{rclcll}
 \coctx{y\!:\!\ident{int}}{ \alift{\cclrd{2}} }
  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~y) &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & (\textit{before}) \\
\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\cclrd{2}\aclrd{\ast}\alift{\cclrd{1},\cclrd{1}}}
  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & (\textit{after}) \\
\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\alift{\cclrd{2},\cclrd{2}}}
  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & (\textit{equivalently})
\end{array}
\end{equation*}
%
The type of the function does not change, because the structural type system associates
the annotation $\cclrd{1}$ with the bound variable $x$ and the substitution does not
affect how the variable $x$ is used.

The other aspect demonstrated in the example is how the coeffect of the substituted
variable affects the free-variable context of the substituted expression. Here, the
original variable $y$ is annotated with $\cclrd{2}$ and we substitute it for an
expression $w+z$ with free variables $w,z$ annotated with $\alift{\cclrd{1}, \cclrd{1}}$.
The substitution applies the operation $\aclrd{\ast}$ (which stands for the sequential
composition $\aseq$ from the structural coeffect algebra) to the annotation of the new context --
in the above example the coeffect $\cclrd{2}\cclrd{\ast}\alift{\cclrd{1}, \cclrd{1}}$ (\emph{after}) is
equivalent to the coeffect $\alift{\cclrd{2}, \cclrd{2}}$ (\emph{equivalently}).

@@

In the Chapter~\ref{ch:flat}, we discussed call-by-value separately from call-by-name, because the
proof of call-by-value substitution has fewer prerequisites. In this section, we consider full
$\beta$-reduction, which encompasses both call-by-value and call-by-name. We also show that
$\eta$-expansion preserves the types. Both of the properties hold for a system with any
structural coeffect algebra that satisfies the additional weak requirements given in Lemma~\ref{thm:structural-substitution}.

>>>>

We mirror the development for flat coeffect calculus and develop a small calculus with a type system
that captures per-variable contextual properties. We outline its categorical semantics and use it as
the basis for a translation that turns well-typed programs in context-aware languages into well-typed
programs in a simple target functional language. We prove syntactic safety for a sample target
language, showing that ``well-typed context-aware programs do not get stuck''.

@@

\item We give a \emph{translational semantics} (Section~\ref{sec:struct-transl}) that translates
  programs from the structural coeffect calculus into a simple functional language with
  uninterpreted comonadically-inspired primitives. We give concrete operational semantics for the
  target language for one of our sample languages and show that well-typed programs, produced by
  translation from the coeffect calculus, do not get stuck.
\end{itemize}

@@

In the previous chapter, we discussed the correspondence between coeffects and effects (and
between comonads and monads). As noted in Section~\ref{sec:applications-strucutre-vec}, the
$\lambda$-calculus is asymmetric in that an expression has multiple inputs (variables in the
context), but just a single result (the resulting value). So, while there is only one notion of
monadic effect system, there are two separate notions of coeffect system -- one that keeps coeffect
annotations per-environment and one that keeps coeffect annotations per-variable.

@@

In the structural coeffect calculus, functions are annotated with a primitive (scalar) coeffect
annotation. A vector of variables forming the free-variable context is annotated with a vector of
coeffect annotations. These annotations differ for various coeffect calculi; their properties are
captured by the definition of \emph{structural coeffect algebra} below. The scalar annotations can
be \eg~integers (how many past values we need) or values of a two-point lattice specifying whether
a variable is live or not. The expressions and types of the structural coeffect calculus are defined
as follows:
%
\begin{equation*}
\begin{array}{rcl}
e &::=& x \sep n \sep \lambda x:\tau.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
\tau &::=& \ident{num} \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
\end{array}
\end{equation*}
%
The expressions and types of structural coeffect calculus are similar to those of the flat coeffect
calculus with two differences. First, we omit the \kvd{let} construct in the core language. In
structural coeffects, let binding can be defined as a derived rule using abstraction and application
(Section~\ref{sec:structural-exts-let}). Second, the coeffect annotations $\cclrd{r}, \cclrd{s}, \cclrd{t}$
on function type now range over values of a \emph{structural coeffect algebra}.

@@

The \emph{structural coeffect algebra} structure is similar to that of \emph{flat coeffect algebra}
with the exception that it drops the $\czip$ operation. It only provides a monoid $(\C, \cseq, \cunit)$
modelling sequential composition of computations and a monoid $(\C, \cpar, \czero)$ representing
pointwise composition, as well as the $\cleq$ relation.

\begin{definition}
\label{def:structural-scalar}
A \emph{\cclrd{structural coeffect algebra}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set
$\C$ together with elements $\cunit, \czero \in \C$, binary operations $\cseq, \cpar$ such that
$(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and a binary relation $\cleq$ such
that $(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
%
\begin{equation*}
\begin{array}{ccr}
r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
\textnormal{(monoid)}
\\
r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
\czero \;\cpar\; r = r = r \;\cpar\; \czero &
\textnormal{(monoid)}
\\
\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
t\; \cleq\; t &
\textnormal{(pre-order)}
\\
\end{array}
\end{equation*}
%
In addition, the following distributivity axioms hold:
\begin{align*}
\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
\end{align*}
\end{definition}

\noindent
The structural coeffect algebra follows the flat coeffect algebra and does not require symmetry of 
$\cpar$ and the annihilation law ($\czero\,\cseq\,\cclrd{r} = \czero$), although both of the properties
hold for the examples in Section~\ref{sec:struct-calculus-examples}. The structural coeffect algebra 
differs from flat coeffect algebra in two important ways:
%
\begin{itemize}
\item The $\cpar$ operation of structural coeffect algebra is not required to be idempotent. In structural
  systems, we can track individual variable accesses and not requiring idempotence allows interesting
  systems such as that for bounded reuse (Section~\ref{sec:struct-semantics-examples}).

@@

\noindent
In the structural coeffect calculus, the scalar coeffect algebra is supplemented by a vector
structure. The vector structure is used to manipulate vectors of coeffect scalars attached to a
variable context. The required structure is captured by the following definition.

\begin{definition}
A \emph{\aclrd{structural coeffect vector}} is formed by a structural coeffect algebra
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ equipped with the following additional structures:

\begin{itemize}
\vspace{-0.3em}
\item Coeffect vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$, ranging over
  elements of the structural coeffect algebra indexed by vector lengths $\sclrd{m}, \sclrd{n} \in \sclrd{\mathbb{N}}$.

@@

In the structural system, this is made explicit by adding \emph{structural rules}.
While substructural type systems usually remove some of the rules, we keep all three and
use them to track how variables are used. This is done by manipulating the coeffect annotations
in parallel with manipulating the variable contexts. As in substructural type systems, (\emph{app})
checks the types of sub-expressions in disjoint parts of the free variable contexts and
(\emph{var}) requires the context to contain exactly one variable. The typing rule for let binding
is a derived rule and we discuss it later in Section~\ref{sec:structural-exts-let}.

@@

\paragraph{Variable contexts.}
In Chapter~\ref{ch:flat}, the free variable context $\Gamma$ was treated as a set. In the type
system for the structural coeffect calculus, the variable context is treated as a vector, with
an additional condition that a variable cannot appear multiple times. We also write $\xlen{-}$ for the
length of the vector:
%
\begin{equation*}
\begin{array}{l}
 \Gamma = \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle \qquad \textnormal{such that}~~\forall i, j ~.~ i \neq j \implies x_i \neq x_j\\
 \xlen{\langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle} = n
\end{array}
\end{equation*}
%
We use the usual notation $x_1\!:\!\tau_1, \ldots, x_1\!:\!\tau_1 \vdash e : \tau$ for
typing judgements, but the free variable context should be understood as a vector.
The notation $\Gamma_1, \Gamma_2$ is used for concatenation of vectors of variables. That is,
given a context $\Gamma_1 = \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle $ and a context
$\Gamma_2 = \langle x_{n+1}\!:\!\tau_{n+1}, \ldots, x_m\!:\!\tau_m \rangle $ then
$\Gamma_1, \Gamma_2 = \langle x_1\!:\!\tau_1, \ldots, x_m\!:\!\tau_m \rangle$.

In the typing rules, free variable contexts are annotated with structural coeffect vectors,
written as $\coctx{x_1\!:\tau_1,\ldots,x_n\!:\!\tau_n}{\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}}$.
Meta-variables ranging over coeffect vectors are written as $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$
(using bold face and colour to distinguish them from scalar meta-variables) and the length of a
coeffect vector is written as $\alen{\aclrd{\textbf{r}}}$.

@@

\paragraph{Structural rules.}
These are shown in Figure~\ref{fig:struct-types} (b). The three structural rules are not
syntax-directed and allow different transformations of the free variable context. They correspond
to the transformations known as \emph{weakening}, \emph{exchange} and \emph{contraction}
from substructural systems.

@@

This suggests that $\czip$ and $\cpar$ serve two roles in flat coeffects. First, they are used
as under-approximations and over-approximations of $\atimes$. This is demonstrated by the
(\emph{approximation}) requirement introduced in Section~\ref{sec:flat-syntax-cbv}, which requires that
$\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$. Semantically, flat abstraction
combines two values representing the available context, potentially discarding parts of it
(under-approximation), while flat application splits the available context (a single value),
potentially duplicating parts of it (over-approxi\-mation)\footnote{Because of this duality, earlier
version of coeffect systems \cite{coeffects-icalp13} used $\wedge$ and $\vee$.}.

@@

\subsection{Examples of structural coeffects}
\label{sec:struct-calculus-examples}

The structural coeffect calculus above can be instantiated to obtain the three structural coeffect calculi
presented in Section~\ref{sec:applications-structural}. Two of them -- structural dataflow and
structural liveness provide a more precise tracking of properties that can be tracked using
flat systems. Formally, any flat coeffect algebra can be turned into a structural coeffect scalar
(by dropping the $\czip$ operator). This is useful for liveness and dataflow, but it does not yield
a practically useful system for the flat algebra coeffect for implicit parameters.

On the other hand, some of the structural systems do not have a flat equivalent, typically
because there is no appropriate $\czip$ operator that could be added to form the flat coeffect
algebra. This is the case, for example, for the system tracking bounded variable use
(Example~\ref{ex:structural-ex-bll}).

\begin{example}[Structural liveness]
The structural coeffect algebra for liveness is formed by
$(\mathcal{L}, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$, where $\mathcal{L}=\{ \ident{L}, \ident{D} \}$ is
the same two-point lattice as in the flat version, that is $\ident{D} \sqsubseteq \ident{L}$
with a join $\sqcup$ and a meet $\sqcap$.
\end{example}

\begin{example}[Structural dataflow]
In dataflow, context is annotated with natural numbers and the structural coeffect algebra is formed
by $(\mathbb{N}, +, \mathit{max}, 0, 0, \leq)$.
\end{example}

\noindent
These two examples have both flat and structural versions. For them, obtaining the structural
coeffect vector is easy. As shown by the examples above, we simply omit the $\czip$ operation. The
laws required by a structural coeffect vector are the same as those required by the flat version
and so the above definitions are both valid. Similar construction can be used for the
\emph{optimized dataflow} example from Section~\ref{sec:flat-calculus-examples}.

It is important to note that this gives us a system with \emph{different} properties.
Information is now tracked per-variable rather than for entire contexts. For dataflow,
we also need to adapt the typing rule for the \kvd{prev} construct. Here, we write $\aclrd{+}$
for a pointwise extension of the $+$ operator, such that
$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} \aclrd{+} k = \alift{\cclrd{r_1} + k, \ldots, \cclrd{r_n} + k}$.
%
\begin{equation*}
\tyrule{prev}
  { \coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau }
  { \coctx{\Gamma}{\aclrd{\textbf{r}} \,\aclrd{+}\, 1} \vdash \kvd{prev}~e : \tau }
\end{equation*}
%
The rule appears similar to the flat one, but there is an important difference. Because of the
structural nature of the type system, it only increments the required number of values for
variables that are actually used in the expression $e$, whereas in the flat coeffect system the rule
incremented the annotation for the whole context. Thanks to the structural nature of the system,
annotations of variables that do not appear in the expression $e$ can be left unchanged.

Before looking at properties of structural coeffect systems,
we consider a system for tracking bounded variable use, which is an example of a structural system
that does not have a flat counterpart.

@@

However, the type system for the structural coeffect calculus in Figure~\ref{fig:struct-types}
introduces another kind of ambiguity due to the fact that non-syntax-directed structural rules
can be applied repeatedly and in arbitrary order. As with the semantics for the flat coeffect calculus
in Chapter~\ref{ch:semantics}, we define the semantics of the structural coeffect calculus relative to
a \emph{typing derivation} and so the meaning of a program depends on the typing derivation chosen.
In this section, we specify how to choose the desired \emph{unique} typing derivation, following
the example of flat coeffect calculi as discussed in Section~\ref{sec:flat-unique}.

@@

As with $\ident{findvar}_{x,\tau}$, the \ident{mergevars} function is fully deterministic.
It returns a context consisting of three parts. Parts $\Gamma_1$
and $\Gamma_2$ represent variables that appear only in the first or the second context; part
$\Gamma$ contains common variables. The coeffect annotations corresponding to $\Gamma_1$ are
the original annotations from $\aclrd{\textbf{r}}$; the coeffects corresponding to $\Gamma_2$
are composed with the coeffect of the function value $\cclrd{t} \,\aseq\, \aclrd{\textbf{s}'}$
as in the original (\emph{app}) rule. Finally, for shared variables, the coeffect is
obtained by point-wise composition (as in contraction) of the coeffect for the two contexts
$\cclrd{r_i} \;\cpar\; (\cclrd{t} \,\cseq\, \cclrd{s_j})$. The first coeffect corresponds to the
context demands in the sub-expression $e_1$ and the second coeffect corresponds to the
function argument $e_2$ sequentially composed with the coeffect $\cclrd{t}$ of the function
(as in ordinary application rule).

@@

When discussing the flat coeffect calculus in Section~\ref{sec:flat-calculus-algebra}, we noted
that the $\cleq$ operation for \emph{flat coeffect algebra} can be defined in terms of $\cpar$
as follows:
%
\begin{equation*}
\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}
\end{equation*}
%
This is not the case for the \emph{structural coeffect algebra} structure. For example, in the
calculus for tracking bounded reuse, the $\cpar$ operator is defined as $+$ (on integers) and
$\cleq$ is just $\leq$ and so the above equivalence does not hold. For this reason, we included
$\cleq$ as an explicit part of both of the structures.

@@

The properties of the structural coeffect algebra, together with two additional weak conditions,
guarantee that certain equational properties on terms hold in all instances of the structural
coeffect calculus that we consider in this thesis. In this section, we look at these common
properties. In Section~\ref{sec:struct-syntax-intro}, we first briefly compare the equational theories
for flat coeffects (Section~\ref{sec:flat-syntax}) and structural coeffects (Section~\ref{sec:struct-syntactic-thms}).

@@

The following example shows a situation similar to the previous one. Here, we use
structural dataflow calculus (writing $\kvd{prev}~e$ to obtain the previous value of the
expression $e$) and we substitute $w+z$ for $y$:
%
\begin{equation*}
\begin{array}{rclcll}
 \coctx{y\!:\!\ident{int}}{ \alift{\cclrd{2}} }
  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~y) &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & (\textit{before}) \\
\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\cclrd{2}\aclrd{\ast}\alift{\cclrd{1},\cclrd{1}}}
  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & (\textit{after}) \\
\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\alift{\cclrd{2},\cclrd{2}}}
  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & (\textit{equivalently})
\end{array}
\end{equation*}
%
The type of the function does not change, because the structural type system associates
the annotation $\cclrd{1}$ with the bound variable $x$ and the substitution does not
affect how the variable $x$ is used.

The other aspect demonstrated in the example is how the coeffect of the substituted
variable affects the free-variable context of the substituted expression. Here, the
original variable $y$ is annotated with $\cclrd{2}$ and we substitute it for an
expression $w+z$ with free variables $w,z$ annotated with $\alift{\cclrd{1}, \cclrd{1}}$.
The substitution applies the operation $\aclrd{\ast}$ (which stands for the sequential
composition $\aseq$ from the structural coeffect vector) to the annotation of the new context --
in the above example the coeffect $\cclrd{2}\cclrd{\ast}\alift{\cclrd{1}, \cclrd{1}}$ (\emph{after}) is
equivalent to the coeffect $\alift{\cclrd{2}, \cclrd{2}}$ (\emph{equivalently}).

@@

In Chapter~\ref{ch:flat}, we discussed call-by-value separately from call-by-name, because the
proof of call-by-value substitution has fewer prerequisites. In this section, we consider full
$\beta$-reduction, which encompasses both call-by-value and call-by-name. We also show that
$\eta$-expansion preserves the types. Both of the properties hold for a system with any
structural coeffect algebra that satisfies the additional weak requirements given in Lemma~\ref{thm:structural-substitution}.
